---
title: "Analysis of Sleep Health and Lifestyle Factors"
author: "Group 20"
date: "December 2nd, 2024"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

-   Zulqarnayeen Sadid (169075017)
-   Amelie Chu Moy San (169076891)

```{r setup, include=FALSE}

# Put any libraries that you need to load here.
# DO NOT PUT "install.packages()" IN AN RMD FILE!!!
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4)
library(tidyverse)
library(lubridate)
library(patchwork)
library(caret)
library(glmnet)
library(tidymodels)
library(stringr)
library(ggridges)
```

# Introduction

Sleep health is a critical component of overall well-being, influencing physical health, mental clarity, and emotional stability. Understanding the various lifestyle factors that affect sleep can provide insights into improving sleep quality across different populations. This analysis aims to explore the relationships between sleep duration, quality of sleep, and various lifestyle factors such as physical activity levels, stress levels, and demographic characteristics.

# Goals/Research Question

The primary goal of this analysis is to investigate how lifestyle factors influence sleep health. Specifically, we aim to answer the following questions:

-   How does physical activity level correlate with sleep duration?
-   What is the relationship between stress levels and quality of sleep?
-   Are there significant differences in sleep patterns based on demographic factors such as gender and age?

We will consider our goal achieved if we can develop a model that predicts sleep efficiency with a root mean square error (RMSE) below 10% on the test set and identify at least three significant factors influencing sleep quality.

# Data Description

The dataset used for this analysis contains information on individuals' sleep duration, quality of sleep, physical activity levels, stress levels, and various demographic factors. The key variables include:

-   **Sleep Duration** (Continuous): Hours of sleep per night.
-   **Quality of Sleep** (Categorical): Rated on a scale from 1 to 10.
-   **Physical Activity Level** (Continuous): Measured in minutes per week.
-   **Stress Level** (Continuous): Self-reported stress level on a scale from 1 to 10.
-   **Gender** (Categorical): Male or Female.
-   **Age** (Continuous): Age of participants in years.

```{r load_and_clean_data}

# Read the data
sleep_data <- read_csv("Sleep_health_and_lifestyle_dataset.csv")

sleep_data
```

## Basic Data Cleaning

```{r load_data}
# Basic cleaning and preprocessing the data
sleep_data_clean <- sleep_data |>
  
  rename_with(~ gsub(" ", "_", .x)) |>

  mutate(
    Sleep_Disorder = factor(Sleep_Disorder, levels = c("None", "Insomnia", "Sleep Apnea")),
    Gender = factor(Gender),
    BMI_Category = factor(BMI_Category),
    Occupation = factor(Occupation),
    Blood_Pressure = str_extract(Blood_Pressure, "\\d+") |> as.numeric(),
    Date = ymd(Sys.Date()) - days(sample(1:365, n(), replace = TRUE))
  ) |>
  
  group_by(Occupation) |>
  mutate(Avg_Stress = mean(Stress_Level)) |>
  ungroup() |>
  mutate(Stress_Relative = Stress_Level - Avg_Stress)

# Custom function to categorize sleep quality
categorize_sleep <- function(duration, quality) {
  case_when(
    duration >= 7 & quality >= 7 ~ "Good",
    duration < 6 | quality < 5 ~ "Poor",
    TRUE ~ "Average"
  )
}


sleep_data_clean <- sleep_data_clean |>
  mutate(Sleep_Category = categorize_sleep(Sleep_Duration, Quality_of_Sleep))



sleep_data_clean
```

The basic data cleaning process involved renaming variables for consistency, converting categorical variables to factors, extracting numeric values from blood pressure readings, and creating derived variables, such as relative stress levels. In order to evaluate sleep quality, based on duration and self-reported quality, we additionally created a custom function.

## Advanced Data Cleaning [Option 1]

```{r advanced_cleaning}

sleep_data_clean <- sleep_data_clean |>
  
  mutate(
    Sleep_Efficiency = Quality_of_Sleep / Sleep_Duration,
    Age_Group = cut(Age, breaks = c(0, 30, 45, 60, Inf),
                    labels = c("Young", "Middle", "Senior", "Elderly")),
    BMI_Numeric = case_when(
      BMI_Category == "Normal" ~ 22,
      BMI_Category == "Overweight" ~ 27,
      BMI_Category == "Obese" ~ 32
    ),
    Sleep_Duration_Zscore = scale(Sleep_Duration),
    Quality_of_Sleep_Zscore = scale(Quality_of_Sleep)
  ) |>
  group_by(BMI_Category) |>
  mutate(
    Sleep_Efficiency_Outlier = abs(scale(Sleep_Efficiency)) > 3,
    Sleep_Efficiency = if_else(Sleep_Efficiency_Outlier, 
                               median(Sleep_Efficiency),
                               Sleep_Efficiency)
  ) |>
  ungroup() |>
  mutate(
    Occupation_Group = fct_lump(Occupation, n = 5),
    Blood_Pressure_Category = case_when(
      Blood_Pressure < 120 ~ "Normal",
      Blood_Pressure < 130 ~ "Elevated",
      Blood_Pressure < 140 ~ "High Stage 1",
      TRUE ~ "High Stage 2"
    )
  )

# Regular expression to extract systolic and diastolic BP
sleep_data_clean <- sleep_data_clean |>
  mutate(
    Systolic_BP = as.numeric(str_extract(Blood_Pressure, "^\\d+")),
    Diastolic_BP = as.numeric(str_extract(Blood_Pressure, "\\d+$"))
  )

# Spliting the data
set.seed(123)
data_split <- initial_split(sleep_data_clean, prop = 0.7, strata = Sleep_Disorder)
train_data <- training(data_split)
temp_data <- testing(data_split)
val_test_split <- initial_split(temp_data, prop = 0.5, strata = Sleep_Disorder)
val_data <- training(val_test_split)
test_data <- testing(val_test_split)

sleep_data_clean
```

In the advanced data cleaning, we created a Sleep Efficiency metric, categorized age into groups, estimated numeric BMI values, standardized sleep duration and quality, handled outliers in sleep efficiency, grouped occupations, categorized blood pressure, and extracted systolic and diastolic blood pressure values using regular expressions & split the data.

# Exploratory Data Analysis

### Exploratory Plot 1: Physical Activity vs. Sleep Duration

```{r insight1, fig.cap="Physical Activity vs. Sleep Duration"}

plot1 <- ggplot(train_data, 
                aes(x = Physical_Activity_Level,
                                y = Sleep_Duration, color = BMI_Category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Gender) +
  labs(title = "Physical Activity vs. Sleep Duration",
       subtitle = "Grouped by BMI Category and Gender",
       x = "Physical Activity Level (minutes/week)",
       y = "Sleep Duration (hours)",
       color = "BMI Category") +
  theme_minimal()

plot1
```

This plot reveals a positive correlation between physical activity levels and sleep duration across different BMI categories and genders. Individuals with higher physical activity tend to have longer sleep durations, with some variations observed between BMI categories and genders.

### Exploratory Plot 2: Stress Level vs. Quality of Sleep

```{r insight2, fig.cap="Stress Level vs. Quality of Sleep"}
plot2 <- ggplot(train_data, 
                aes(x = Stress_Level, 
                    y = Occupation_Group, fill = Quality_of_Sleep)) +
  geom_density_ridges(alpha = 0.7, scale = 0.9) +
  labs(title = "Stress Level vs. Quality of Sleep",
       subtitle = "Distribution by Occupation Group",
       x = "Stress Level",
       y = "Occupation Group",
       fill = "Quality of Sleep") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot2
```

This plot illustrates the distribution of sleep quality across different stress levels & occupations. It suggests that higher stress levels are generally associated with lower sleep quality, with variations observed across different occupations.

# Modeling

### Linear Model for Relationship between Sleep Duration & Stress and Physical Activity:

```{r linear_model_1}
model_plot1 <- ggplot(train_data,
                      aes(x = Stress_Level,
                          y = Sleep_Duration, 
                          color = Physical_Activity_Level)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Sleep_Disorder) +
  scale_color_viridis_c() +
  labs(title = "Sleep Duration vs. Stress and Physical Activity",
       subtitle = "Grouped by Sleep Disorder",
       x = "Stress Level",
       y = "Sleep Duration (hours)",
       color = "Physical Activity\n(minutes/week)") +
  theme_minimal()

model_plot1
```

This plot shows the relationship between stress levels, physical activity, and sleep duration, categorized by sleep disorders. It suggests that higher stress levels are associated with shorter sleep duration, while higher physical activity levels tend to correlate with longer sleep duration.

### Linear Model for Relationship between Sleep Quality and Age & BMI

```{r linear_model_2}
model_plot2 <- ggplot(train_data,
                      aes(x = Age, 
                          y = Quality_of_Sleep,
                          color = BMI_Category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~Gender) +
  labs(title = "Sleep Quality vs. Age and BMI",
       subtitle = "Grouped by Gender",
       x = "Age",
       y = "Quality of Sleep",
       color = "BMI Category") +
  theme_minimal()

model_plot2
```

This plot illustrates the relationship between age, BMI category, and sleep quality, separated by gender. It reveals potential age-related trends in sleep quality and differences between BMI categories and genders.

## Exploratory Linear Model

### Exploratory Linear Model 1

```{r exploratory_linear_model_1}
model1 <- lm(Sleep_Duration ~ Stress_Level + 
               Physical_Activity_Level + 
               BMI_Category + 
               Gender + 
               Age, data = train_data)
summary(model1)

# Calculate RMSE on validation set
predictions1 <- predict(model1, newdata = val_data)
rmse1 <- sqrt(mean((val_data$Sleep_Duration - predictions1)^2))
cat("RMSE for Model 1:", rmse1, "\n")

# Try removing Gender and adding an interaction term
model1_mod <- lm(Sleep_Duration ~ 
                   Stress_Level * Physical_Activity_Level + 
                   BMI_Category + Age, data = train_data)
predictions1_mod <- predict(model1_mod, newdata = val_data)
rmse1_mod <- sqrt(mean((val_data$Sleep_Duration - predictions1_mod)^2))
cat("RMSE for Modified Model 1:", rmse1_mod, "\n")
```

This exploratory model examines factors influencing sleep duration. The modification, which includes an interaction term between stress level and physical activity, slightly improves the model's performance on the validation set.

### Exploratory Linear Model 2

```{r exploratory_linear_model_2}
model2 <- lm(Quality_of_Sleep ~ Age +
               Heart_Rate + 
               Blood_Pressure + 
               Occupation + 
               Sleep_Duration, data = train_data)
summary(model2)

# Calculate RMSE on validation set
predictions2 <- predict(model2, newdata = val_data)
rmse2 <- sqrt(mean((val_data$Quality_of_Sleep - predictions2)^2))
cat("RMSE for Model 2:", rmse2, "\n")

# Try adding an interaction term and removing Occupation
model2_mod <- lm(Quality_of_Sleep ~ Age * 
                   Sleep_Duration + 
                   Heart_Rate + 
                   Blood_Pressure, data = train_data)
predictions2_mod <- predict(model2_mod, newdata = val_data)
rmse2_mod <- sqrt(mean((val_data$Quality_of_Sleep - predictions2_mod)^2))
cat("RMSE for Modified Model 2:", rmse2_mod, "\n")
```

This model explores factors influencing sleep quality. The modification, which includes an interaction between age and sleep duration, shows a slight improvement in predictive performance on the validation set.

## Final Linear Model

```{r final_linear_model}
# Combine training and validation sets
train_val_data <- bind_rows(train_data, val_data)

# Train final models
final_model1 <- lm(Sleep_Duration ~ Stress_Level *
                     Physical_Activity_Level + 
                     BMI_Category + 
                     Age, data = train_val_data)
final_model2 <- lm(Quality_of_Sleep ~ Age * 
                     Sleep_Duration + 
                     Heart_Rate + 
                     Blood_Pressure, data = train_val_data)

# Make predictions on test set
test_pred1 <- predict(final_model1, newdata = test_data)
test_pred2 <- predict(final_model2, newdata = test_data)

# Calculate RMSE on test set
test_rmse1 <- sqrt(mean((test_data$Sleep_Duration - test_pred1)^2))
test_rmse2 <- sqrt(mean((test_data$Quality_of_Sleep - test_pred2)^2))

cat("Test RMSE for Final Model 1 (Sleep Duration):", test_rmse1, "\n")
cat("Test RMSE for Final Model 2 (Sleep Quality):", test_rmse2, "\n")
```

The final linear models, trained on the combined training and validation sets, show reasonable performance on the test set for both sleep duration and quality predictions.

# Conclusion

Our study discovered a number of important factors which influence sleep disorders, such as age, occupation, BMI, physical activity, and stress levels. The created linear models demonstrate favorable effects in terms of forecasting the duration and quality of sleep, and in certain circumstances, including interactive aspects improves the model performance.

# Limitations

**Data**: The dataset may not be fully representative of the general population, and self-reported data could introduce bias. The cross-sectional nature of the data limits our ability to infer causal relationships.

**Modelling**: Linear models may not capture complex, non-linear relationships between variables. The simplification of sleep disorders into three categories may overlook nuances in sleep health.

# References

Further insights were derived from:

\- Buysse, D. J. (2014)[^1].

[^1]: Buysse, D. J. (2014). Sleep health: can we define it? Does it matter?. Sleep, 37(1), 9-17. url: <https://academic.oup.com/sleep/article-abstract/37/1/9/2454038>

\- Grandner, M. A., & Malhotra, A. (2015)[^2].

[^2]: Buysse, D. J. (2014). Sleep health: can we define it? Does it matter?. Sleep, 37(1), 9-17. url: <https://academic.oup.com/sleep/article-abstract/37/1/9/2454038>

\- Knutson, K. L., et al. (2017)[^3].

[^3]: Grandner, M. A., & Malhotra, A. (2015). Sleep as a vital sign: why medical practitioners need to routinely ask their patients about sleep. Sleep Health, 1(1), 11-14. url: <https://pmc.ncbi.nlm.nih.gov/articles/PMC5102393/>

\- Matricciani, L., et al. (2017)[^4].

[^4]: Knutson, K. L., et al. (2017). The National Sleep Foundation's sleep health index. Sleep Health, 3(4), 234-240. url: <https://pubmed.ncbi.nlm.nih.gov/28923186/>

\- Medic, G., et al. (2017)[^5].

[^5]: Matricciani, L., et al. (2017). Past, present, and future: trends in sleep duration and implications for public health. Sleep Health, 3(5), 317-323. url: <https://pubmed.ncbi.nlm.nih.gov/28923186/>

\- Ohayon, M., et al. (2017)[^6].

[^6]: Medic, G., et al. (2017). Short- and long-term health consequences of sleep disruption. Nature and Science of Sleep, 9, 151-161. url: <https://pmc.ncbi.nlm.nih.gov/articles/PMC5449130/>

\- Wickham, H., et al. (2019)[^7].

[^7]: Ohayon, M., et al. (2017). National Sleep Foundation's sleep quality recommendations: first report. Sleep Health, 3(1), 6-19. url: <https://escholarship.org/uc/item/9xc5x5h2>

\- Kuhn, M., & Wickham, H. (2020)[^8].

[^8]: Wickham, H., et al. (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686. url: <https://joss.theoj.org/papers/10.21105/joss.01686>
